Reconstructing Propositional Proofs in Type Theory

### 2017-11-20 paper/sections/type-theory.tex
[ paper ] lots of changes.

Type theory is a formalism for the foundation of mathematics, and it has became in a key instrument to study logic and proof theory.
It was proposed in 1902 to avoid Russell's paradox and it follows
the same philosophy of constructive mathematics where the witness of
a statement is \emph{everything}. In that direction, we could say
the main actors in type theory are the statements and their proofs.
By following the Curry-Howard correspondence (see, for example,~\cite{Wadler-2015}), types as propositions.
A proposition corresponds to a type, and a proof of that propositions
corresponds to the term of that proof (henceforth proof-term).
%\emph{types} are theorems and their \emph{terms} are their proofs.

Since type theory is a formal system, we have a syntax and a set of
derivation rules. These rules enables us to derive a kind of conclusions called
\emph{judgments}. A judgment is another way to say that a term has a certain type.

@@

Therefore, the
proof verification task becomes in type theory as verifying that the proof-term has the correspondent type of the theorem. We know
this process as \emph{type-checking}. This feature of type theory allow us to
verify a proof generated by an \ATP by reconstructing its proof-term to type-
check the proof. For such a purpose, we use a proof assistant based on type
theory like \Agda to delegate this task.
% However, this approach has some weak points discussed in~\cite{Bezem2002}.
Some limitations from the type theory
point of view for proof-reconstruction is described in the following sections.

@@

To reconstruct \Metis inference rules in type theory, we observed that some
rules or their inner functions are implemented by \emph{general recursive}
functions.

Functions defined by a general recursion can not be directly translated in type
theory since it is not a guarantee they terminate. For that reason, we follow
the technique described in~\cite{Bertot2004} to avoid termination problems by
modifying the recursive functions to be \emph{structurally recursive}. % (see
\cite{Coquand1992,Abel2002,Bove2005} for more details).

This paper adopts the notion of \emph{structurally smaller}
from~\cite{Abel2002}, that is, a recursive function is structurally
recursive if it calls itself with only structurally smaller arguments.

General recursive functions can be rewrote into structurally recursive
functions by using for instance, the \emph{bounded recursion} technique.
For a yet another methods, we refer
the reader to \cite{Coquand1992,Abel2002,Bove2005}.
The bounded technique defines a new structural recursive function based on
the general recursive function by adding an argument.
The new argument is the \emph{bound}, a natural number given by the
function complexity. In other words, the added argument will represent
the number of times the function needs to call itself to get the
expected outcome.

\begin{notation}
  We use \Prop for the type of propositions.
  We write the equality as $φ ≡ ψ$, for $φ,~ψ :
  \Prop$. For this equality, we assume the reflexivity, symmetry and
  transitivity properties.
  We say $\varphi$ and $\psi$ are Two propositions are syntactically equivalent if there is a derivation from $\varphi$ to $\psi$ and
  a derivation from $\psi$ to $\varphi$.
   The type of natural numbers is called \Nat,
  and it is defined as usual, where $\zero$ and $\suc$ are its data
  constructors. We use names and symbols for the arithmetic operations
  as usual (\eg, $+$, $-$, $*$).  We use syntax sugar for $\zero$, $\suc$,
  $\suc~\zero$, $\dots$, with the decimal representation~$0$, $1$,
  $2$, $\dots$ as well.
\end{notation}

Hence, one approach to define a structural recursive function based on a general
recursive function $\rm{f}:~\abbre{a}~\to~\abbre{b}$ is to formulate a new
function $\rm{f}^{*}~:~\abbre{a}~\to~\Nat~\to~\abbre{b}$ where all of its
recursive calls are done on smaller arguments.

@@

Let us consider the following example to show the bounded recursion technique
for the \fuh function. This function is used for reconstructing
a \Metis inference rule in Section~\ref{sssec:strip-a-goal}.

@@

The complexity measure of a function is the number of steps
the function takes to finish. Since the functions defined for formulas
are based on the formula structure, we assign a complexity measure for each
of these cases in the formula structure. We refer the reader to~\cite{VanDalen1994,Agudelo-Agudelo2017} for more details about this kind of function.

In our case, we define the $\fuh_{cm}: \Prop → \Nat$ function as
the complexity measure for the $\fuh_0$ function.

@@

For a more complete treatment of general recursion, and how to define
well-founded and well-structural recursive functions, we refer the
reader to~\cite{Bove2005}.

@@

Type theory and intuitionistic logic are accompanying theories since
they are based on the same basis of constructive mathematics. Proving
propositions in these theories demands a witness construction for the
proof. Nonetheless, classical logic does not always have constructive
proofs since some of the proofs are stated by refutation.

@@

\begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ ψ$}
    \RightLabel{$\vee$-intro$_2$}
    \UnaryInfC{$Γ ⊢ φ \vee ψ$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ , φ ⊢ γ $}
    \AxiomC{$Γ , ψ  ⊢ γ$}
    \RightLabel{$\vee$-elim}
    \BinaryInfC{$Γ , φ \vee ψ ⊢ γ$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ , φ ⊢ ψ$}
    \RightLabel{$\Rightarrow$-intro}
    \UnaryInfC{$Γ ⊢ φ \Rightarrow ψ$}
  \end{bprooftree}
\end{equation*}

\[
%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ \Rightarrow ψ$}
\AxiomC{$Γ ⊢ φ$}
\RightLabel{$\Rightarrow$-elim}
\BinaryInfC{$Γ ⊢ ψ$}
\end{bprooftree}\qquad
\begin{bprooftree}
\AxiomC{\,}
\RightLabel{PEM}
\UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$}
\end{bprooftree}
\]

>>>>

% It was proposed in 1902 to avoid Russell's paradox --- (bettter mention BHK)
Type theory is a formalism for the foundation of mathematics,
it has became in a key instrument to study logic and proof theory
that follows the same basis of constructive mathematics where the witness of a statement is \emph{everything}.
In that direction, we could say the main actors in type theory are the statements and their proofs.

By following the Curry-Howard correspondence (see, for
example,~\cite{Wadler-2015}), types are propositions. The inhabitants
of that types, the \emph{terms}, are proofs of that propositions.
Since type theory is a formal system, we have a syntax and a set of
derivation rules. These rules enables us to derive a kind of
conclusions called \emph{judgments}. A judgment is another way to say
that a term has a certain type.

@@

Therefore, the proof verification task becomes in type theory as
verifying that the proof-term has the correspondent type of the
theorem. We know this process as \emph{type-checking}.
This feature of type theory allow us to verify a proof generated by an \ATP by reconstructing its proof-term to type-check the proof. For such a
purpose, we use a proof assistant based on type theory like \Agda to
delegate this task. % However, this approach has some weak points
discussed in~\cite{Bezem2002}. Some limitations from the type theory
point of view for proof-reconstruction is described in the following
sections.

@@

To reconstruct \Metis inference rules in type theory, we observed
that some rules or their inner functions are implemented by \emph{
general recursive} functions.

Functions defined by a general recursion can not be directly
translated in type theory since it is not a guarantee they terminate.
For that reason, we follow the technique described
in~\cite{Bertot2004} to avoid termination problems by modifying the
recursive functions to be \emph{structurally recursive}.

A recursive function is structurally recursive if it calls itself
with only structurally smaller arguments~\cite{Abel2002}.
General recursive functions can be rewrote into structurally
recursive functions by  using for instance, the \emph{bounded
recursion} technique. For a yet another methods, we refer the reader
to \cite{Coquand1992,Abel2002,Bove2005}.

The bounded technique defines a new structural recursive function
based on the general recursive function by adding an argument. The
new argument is the \emph{bound}, a natural number given by the
function complexity.
In other words, the added argument will represent the number of times
the function needs to call itself to get the expected outcome.

\begin{notation}
  We use \Prop for the type of propositions.
  A proposition is an expression of indivisible propositional variables (\eg, symbols $φ₀, φ₁, \cdots$), the logic constants:
  $⊥$, $⊤$, the binary connectives ($∧, ∨, ⇒, ⇔)$, and the negation
  $(\neg)$.
  We write the equality using the symbol (≡) as $φ ≡ ψ$, for $φ,~ψ :
  \Prop$. For this equality, we assume the reflexivity, symmetry and
  transitivity properties.
  % We say $φ$ and $ψ$ are syntactically equivalent if there is a derivation from $φ$ to $ψ$ and
  % a derivation from $ψ$ to $φ$.
  The type of natural numbers is called \Nat,
  and it is defined as usual, where $\zero$ and $\suc$ are its data
  constructors. We use names and symbols for the arithmetic operations
  as usual (\eg, $+$, $-$, $*$).  We use syntax sugar for $\zero$, $\suc$,
  $\suc~\zero$, $\dots$, with the decimal representation~$0$, $1$,
  $2$, $\dots$ as well.
\end{notation}

Hence, one approach to define a structural recursive function based
on a general
recursive function $\rm{f}:\,\abbre{a}~\to~\abbre{b}$ is to
formulate a new
function $\rm{f}^{*}:\,\abbre{a}~\to~\Nat~\to~\abbre{b}$ where all
of its recursive calls are done on smaller arguments.

\begin{notation}
We define some functions by pattern-matching on \Prop or \Nat.
\end{notation}

@@

Let us consider the following example to show the bounded recursion
technique for defining the \fuh function. This function is used for
reconstructing a \Metis inference rule in
Section~\ref{sssec:strip-a-goal}.

@@

The complexity measure of a function is the number of steps the
function takes to finish. Since the $\fuh_{0}$ function is recursive,
we can define its complexity measure by defining a recursive function
instead of a closed formula for such a number. However, this function
must to be structurally recursive as well and can be defined by
following the pattern-matching cases of its definition.

Therefore, we define in~\eqref{eq:uh-complexity} the complexity
measure function of $\fuh_{0}$ by assigning the number of steps to
finish for each pattern-matching case.
We have used the complexity measure definition for a formula defined in~\cite{Agudelo-Agudelo2017} to define other complexity measures
in this paper.

@@

% For a more complete treatment of general recursion, and how to define
% well-founded and well-structural recursive functions, we refer the
% reader to~\cite{Bove2005}.

@@

Type theory and intuitionistic logic are accompanying theories since
they are based on the same philosophical basis of constructive
mathematics. Proving propositions in these theories demands a witness
construction for the proof. Nonetheless, classical logic does not
always have constructive proofs since some of the proofs are stated
by refutation.

@@

\begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ ψ$}
    \RightLabel{$\vee$-intro$_2$}
    \UnaryInfC{$Γ ⊢ φ \vee ψ$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ , φ ⊢ γ $}
    \AxiomC{$Γ , ψ  ⊢ γ$}
    \RightLabel{$\vee$-elim}
    \BinaryInfC{$Γ , φ \vee ψ ⊢ γ$}
  \end{bprooftree}\quad
  \begin{bprooftree}
    \AxiomC{$Γ , φ ⊢ ψ$}
    \RightLabel{$⇒$-intro}
    \UnaryInfC{$Γ ⊢ φ ⇒ ψ$}
  \end{bprooftree}
\end{equation*}

\[
%\scalebox{0.9}{
\begin{bprooftree}
\AxiomC{$Γ ⊢ φ ⇒ ψ$}
\AxiomC{$Γ ⊢ φ$}
\RightLabel{$⇒$-elim}
\BinaryInfC{$Γ ⊢ ψ$}
\end{bprooftree}\qquad
\begin{bprooftree}
\AxiomC{\,}
\RightLabel{PEM}
\UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$}
\end{bprooftree}
\]
